---
title: "Assignment 8: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Lex Clay" 
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**

>NOTE: For this assignment, you can skip tests for any of the assumptions required to run the models. 

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

#install.packages("agricolae")

library(tidyverse)
library(agricolae)
library(here)
library(corrplot)

#2

NTL.LTR.Raw <- read.csv(here("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Depth across all lakes has no effect on lake temperature recorded in July. 
Ha: Lake temperature recorded in July decreases with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.



```{r}
#Extract Date to get July 

NTL.LTR.Raw$sampledate = ymd(NTL.LTR.Raw$sampledate)
NTL.LTR.Raw$month <- month(NTL.LTR.Raw$sampledate) 
  

```

```{r scatterplot}
#4

#wrangling the raw data set and then applying lm()

NTL.LTR.wrangled <- NTL.LTR.Raw %>% 
  select(lakename, year4, daynum, depth, temperature_C, month) %>% 
  filter(month==07) %>% 
  na.omit()

#5

July.NTL.LTR.scatter <- ggplot(NTL.LTR.wrangled, aes(x=depth, y=temperature_C))+labs(
  x="Depth", 
  y="Temperature"
)+
  ylim(0,35)+
  theme(
    panel.background = element_rect(fill = "#002020"),
    panel.grid = element_blank()
  )+
  geom_smooth(method="lm", se=FALSE, color="purple")+
  geom_point(aes(color=lakename), alpha=0.3, size=1.3)

print(July.NTL.LTR.scatter)


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure(s) suggest(s) that there is a inverse relationship between depth and temperature. As depth increases, temperature tends to decrease. The intercept expects that when the value of the depth is 0 - so at the surface - the temperature will be 21.5 degrees celsius. 

7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

July.Regression.NTL.LTR <- lm(data=NTL.LTR.wrangled, temperature_C ~ depth) 

summary(July.Regression.NTL.LTR)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The slope indicates that with each depth increase of 1, the temperature decreases by -1.9. The P value is 2.2e-16, which is basically 0 == statistically significant. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

#the smaller the AIC value, the better the model performance is. AIC is run by default 
#using the function step(). The best model uses as few predictors as possible. 

#running the linear regression
temp.predictors <- lm(data=NTL.LTR.wrangled, temperature_C~year4+daynum+depth)

#run AIC 
my.aic <-step(temp.predictors)

#10

#running the multiple linear regression  w/ depth and daynum because AIC sky rockets when either of those are removed. Meanwhile year4 AIC is the same as <none> so it has no stat. influence. 

good.model <- lm(data=NTL.LTR.wrangled, temperature_C~daynum+depth)

summary(good.model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: I ran the the linear regression above  w/ depth and daynum because AIC sky rockets when either of those are removed. Meanwhile year4 AIC is the same as <none> so it has no stat. influence. This model - good.model - has a higher R-squared value than the model using only depth (.778 vs .743) and therefore is a better indicator of temperature variance. 
---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

#Wrangle
lake.temps <- NTL.LTR.wrangled %>% 
  group_by(lakename) %>% 
  summarise(average_temp=mean(temperature_C))

summary(lake.temps) #this shows the average temps for each lake but now we need to use anova/lm to determine if those differences are significant. 

anova.temp <- aov(data=NTL.LTR.wrangled, temperature_C ~ lakename)

lm.model <- lm(data=NTL.LTR.wrangled, temperature_C ~ lakename)

summary(anova.temp)
summary(lm.model)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: A low F-value of 1.098 indicates that the difference in mean temperature among the lakes varies about as much as the temperature within the lakes and the the P value .357 suggests that temperature variance between lakes is likely random. The R-squared value indicates that the lake name only accounts for about 1% of temperature reading. Ergo, not statiscally significant. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14


NTL.LTR.faceted <- ggplot(NTL.LTR.wrangled, aes(x=depth, y=temperature_C, color=lakename))+
  geom_point(alpha=0.5, size=1)+ 
  geom_smooth(method="lm", se=FALSE)+
  ylim(0,35)+
  facet_wrap(~lakename)+
   theme(
    panel.background = element_rect(fill = "#002020"),
      panel.grid = element_blank()
   )

#still need to remove panel grid here 
print(NTL.LTR.faceted) 

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(anova.temp)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Since all lakes have a P value far above 0.05, none have a statistically distinct mean temperature. The lakes are pretty much indistinguishable via their temperature data. 
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A T-test. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

#Crampton does not have any July data which is why it was dropped by na.omit() from the raw dataset. I've used Tuesday Lake instead. 

ward.tuesday <- NTL.LTR.wrangled %>% 
  filter(lakename %in% c("Ward Lake","Tuesday Lake"))

ward.tuesday.test <- t.test(data=ward.tuesday, temperature_C ~lakename)

ward.tuesday.test


#mean in group Tuesday Lake    mean in group Ward Lake 
#                 10.11667                   13.93103 


# Ward Lake-Tuesday Lake       3.8143678 
```

>Answer: The Tukey test says the diff between Ward Lake-Tuesday Lake is 3.814, which matches the t.test results to the T. (Sorry, I had to.)
