---
title: "Assignment 5: Data Visualization"
author: "Lex Clay" 
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

library(tidyverse)
library(here)
library(cowplot)

#2 
getwd()
here()

NTL.LTR <-read.csv(file=here('Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv'), stringsAsFactors = TRUE)

NEON.NIWO <-read.csv(file=here('Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv'), stringsAsFactors = TRUE)

str(NTL.LTR)
str(NEON.NIWO)

NTL.LTR$sampledate <-ymd(NTL.LTR$sampledate)

NTL.LTR$sampledate <- as.Date(NTL.LTR$sampledate, format = "%Y-%m-%d")
NEON.NIWO$collectDate <-as.Date(NEON.NIWO$collectDate, format ="%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3


myfirsttheme <- theme_classic(base_size = 12)+
  theme(
    axis.text = element_text(family ="Palatino", color="#0A1A2F"),
        legend.position="top",
  plot.background = element_rect(fill="#FAFAFA", color=NA), plot.title = element_text(color="#0A1A2F"))
      
theme_set(myfirsttheme) #setting this theme as default
   
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

PPO4_Plot <- ggplot(NTL.LTR, aes(x=tp_ug, y=po4, color=lakename)) + 
  geom_point()+
   geom_smooth(method='lm')+
   ylim(0,50)+ 
  labs(
    color="Lake Name", 
    x="Phosphorus",
    y="Phosphate"
  )+
  scale_color_manual(values = c(
    "Paul Lake" = "#1C3144",
    "Peter Lake" = "#6B7F82"
  ))
 
print(PPO4_Plot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. Show all months, even those where no data was collected. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis text in your theme to `element_blank()` removes the axis text (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different relative sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

temp_boxlegend <- ggplot(NTL.LTR, aes(
  x =factor(month),
  y=temperature_C, 
  fill=lakename), 
  )+ 
 labs(
   fill="Lake Name",  #this is where we change the legend title since we extract here
   title ="Monthly Temp",
   x="Months", 
   y="Temperature (Celsius)")+
  theme(
  plot.title =  element_text(margin = margin(b=3))
  )+
  geom_boxplot()

print(temp_boxlegend)


temp_box <- ggplot(NTL.LTR, aes(
  x =factor(month),
  y=temperature_C, 
  fill=lakename), 
  )+ 
 labs(
   title ="Monthly Temp",
   x="Months", 
   y="Temperature (Celsius)")+
  theme(
  legend.position = "none", 
  plot.title =  element_text(margin = margin(b=3))
  )+
  geom_boxplot()
  
print(temp_box)


#trying out with sampledate instead of month factor.ignore this. it's just me experimenting.




#ignore 




tempboxsampledate <-ggplot(NTL.LTR, aes(
  x=factor(format(NTL.LTR$sampledate, "%Y")),
  y=temperature_C)
)+
  geom_boxplot()+
  labs(
    title = "Temp Data", 
    x="date", 
    y="temperature"
  )
  
#print(tempboxsampledate)





```

```{r}
tp_box <-ggplot(NTL.LTR, aes(
  x=factor(month),
  y=tp_ug, 
  fill=lakename)
  )+
labs(
  title = "Phosphorous Totals per Month",
  x="Months",
  y="TP")+ 
theme(
    legend.position = "none",
    plot.title = element_text(margin=margin(b=3))
  )+ 
geom_boxplot()

print(tp_box)

```

```{r}


tn_box <-ggplot(NTL.LTR, aes(
  x=factor(month),
  y=tn_ug, 
  fill=lakename)
  )+
labs(
   title = "Nitrogen Totals per Month",
   x="Months",
   y="TN"
   )+
  theme(
    legend.position ="none", 
    plot.title = element_text(margin = (margin(b=3)))
  )+
  geom_boxplot()
print(tn_box)

```

```{R}

#save legend as object 

grid_legend <- get_legend(
  temp_boxlegend +
    theme(legend.position = "right")
)
```




```{r Combing boxplots}

box_grid_aligned <- plot_grid(temp_box, tp_box, tn_box,align ='hv', rel_widths = c(1,1,1), ncol = 1)
                              
print(box_grid_aligned)


#reference link https://wilkelab.org/cowplot/articles/shared_legends.html


#add legend

box_grid_legendary <- plot_grid(box_grid_aligned, grid_legend, nrow = 1)

print(box_grid_legendary)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Don't know if I am using the wrong dataset considering all nitrogen and phosphorous data is missing a date, so there is not much to observe there other than that Peter Lake has higher totals of both and that temperatures are predictably higher in July and August. 


6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

suppressWarnings({ #open warning suppression

  
#6-a practicing via data wrangling method. see 6-b for subset method. 
niwot.needles <-NEON.NIWO %>% 
  filter(functionalGroup=="Needles")

niwot.needle.box <- ggplot(niwot.needles,aes(
  x= factor(collectDate),
  y=dryMass, 
  fill=nlcdClass, 
  ))+
  labs(
    title = "Needle Dry Mass", 
     x = "Date", 
    y= "Dry Mass",
    fill="NCLD Class"
)+
  geom_boxplot()


print(niwot.needle.box)

#6-b subset method 

NEON.NIWO$Year <- format(NEON.NIWO$collectDate, "%Y")
NEON.NIWO$Month <-format(NEON.NIWO$collectDate, "%m")
  
niwot.subset.box <- ggplot(subset(NEON.NIWO, functionalGroup == "Needles"),
aes(
  x=factor(Month), 
  y=dryMass, 
  fill=nlcdClass
))+
  facet_grid(nlcdClass ~ Year, 
      labeller = labeller(nlcdClass=c("evergreenForest" = "Evergreen Forest",
"grasslandHerbaceous" = "Grassland","shrubScrub"="Shrub Scrub")
))+
  geom_boxplot()+
  labs(
    x="Collection Month", 
    y="Dry Mass", 
    fill="NLCD Class"
)+
  theme(
    axis.text.x = element_text(angle=45, hjust = 1, size = 8)
  )+
scale_fill_discrete(
    labels =c(
      "evergreenForest" = "Evergreen Forest",
      "grasslandHerbaceous" = "Grassland",
      "shrubScrub"="Shrub Scrub")
    )
 

print(niwot.subset.box) #this is the best plot of the two strategies. 


}) #close warning suppression


needles.face <- ggplot(niwot.needles, aes(
  x=factor(collectDate),
  y=dryMass, 
  fill=nlcdClass
  ))+
  labs(
    x="Collection Date",
    y= "Dry Mass", 
    fill= "NLCD Class"
  )+theme(
    axis.text.x = element_text(angle=45, hjust = 1, size=8)
  )+
  geom_boxplot()+
   facet_wrap(vars(nlcdClass),nrow=1)+
  scale_fill_discrete(
    labels =c(
      "evergreenForest" = "Evergreen Forest",
      "grasslandHerbaceous" = "Grassland",
      "shrubScrub"="Shrub Scrub"
    ))

print(needles.face)





```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: The three-faceted plot is more effective. You can easily compare Dry Mass between months and compare between year years. Unlike the other plot, the facet_grid method makes the pattern of decreasing dry mass between 2016 - 2019 obvious at a glance. 
